---
title: Millennial Spending
author: Clayton Knappenberger
date: '2019-05-24'
slug: millennial-spending
categories:
  - consumption
tags: []
---

# Sankey Diagrams
[Sankey Diagrams](https://en.wikipedia.org/wiki/Sankey_diagram) are a popular way of showing flows. Lately they've become pretty popular for showing budgets since they allow you to show the hierarchical relationships and proportions within groups pretty intuitively. Below, I just wanted to provide a really quick example of one of these using data from the [Consumer Expenditure Surveys](www.bls.gov/cex). Speifically, the data used come from the [2017 Generational table](https://www.bls.gov/cex/2017/combined/gener.pdf) to show how the average Millennial spends their money. 

There are a couple of caveats: 

1. Alcohol was combined with food to make a higher-level food and drinks category
2. Reading was moved into the entertainment category
3. Several smaller categories were regrouped into a larger miscellaneous category

Other than that the data match exactly what's in the official table. Enjoy!

```{r, echo=FALSE}
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(networkD3))
suppressPackageStartupMessages(library(htmlwidgets))
suppressPackageStartupMessages(library(htmltools))
```

```{r, echo=FALSE}
nodes = data.frame(name = c("Total", "Food/drink", "At home", "Away from home", "Alcohol", 
                            "Housing", "Shelter", "Utilities", "Operations", "Supplies", "Furnishings", 
                            "Apparel", "Men/boys", "Women/girls", "Infants", "Footwear", "Other", 
                            "Transportation", "Vehicle purchase", "Gas/fuels", "Public transit", "Other", 
                            "Healthcare", "Insurance", "Serivces", "Drugs", "Supplies", 
                            "Entertainment", "Fees/admissions", "Audio/visual equipment/services", "Pets/toys/hobbies",
                            "Reading", "Other",
                            "Misc", "Personal care products/services", "Education", "Tobacco/smoking products",
                            "Contributions", "Other", 
                            "Personal Insurance/pensions", "Insurance", "Social Security/pensions"), 
                   group = c("Total", "Food/drink", "Food/drink", "Food/drink", "Food/drink", 
                             "Housing", "Housing", "Housing", "Housing", "Housing", "Housing",
                             "Apparel", "Apparel", "Apparel", "Apparel", "Apparel", "Apparel", 
                             "Transportation", "Transportation", "Transportation", "Transportation", "Transportation", 
                             "Healthcare", "Healthcare", "Healthcare", "Healthcare", "Healthcare", 
                             "Entertainment", "Entertainment", "Entertainment", "Entertainment", "Entertainment",
                             "Entertainment", 
                             "Misc", "Misc", "Misc", "Misc", "Misc", "Misc", 
                             "Personal Insurance/pensions", "Personal Insurance/pensions", 
                             "Personal Insurance/pensions"))

links = data.frame(source = c(0, 1, 1, 1, 0, 5, 5, 5, 5, 5, 0, 11, 11, 11, 11, 11, 0, 17, 17, 17, 17, 
                              0, 22, 22, 22, 22, 0, 27, 27, 27, 27, 27, 0, 33, 33, 33, 33, 33, 0, 39, 39), 
                   target = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 
                              22, 23, 24, 25, 26, 27, 28, 29, 30, 31, 32, 33, 34, 35, 36, 37, 38, 39, 40, 41), 
                   value = c(7424, 3680, 3223, 521, 18309, 11566, 2996, 1523, 495, 1729,
                             1908, 443, 616, 152, 410, 287, 8629, 3825, 1919, 610, 2275, 
                             2894, 2043, 565, 201, 85, 2559, 579, 782, 629, 63, 506, 
                             3863, 616, 1457, 303, 842, 645, 5720, 172, 5548), 
                   group = c("Food/drink", "Food/drink", "Food/drink", "Food/drink", 
                             "Housing", "Housing", "Housing", "Housing", "Housing", "Housing",
                             "Apparel", "Apparel", "Apparel", "Apparel", "Apparel", "Apparel", 
                             "Transportation", "Transportation", "Transportation", "Transportation", "Transportation", 
                             "Healthcare", "Healthcare", "Healthcare", "Healthcare", "Healthcare", 
                             "Entertainment", "Entertainment", "Entertainment", "Entertainment", "Entertainment",
                             "Entertainment", 
                             "Misc", "Misc", "Misc", "Misc", "Misc", "Misc", 
                             "Personal Insurance/pensions", "Personal Insurance/pensions", 
                             "Personal Insurance/pensions"))

sankeyNetwork(Links = links, Nodes = nodes, Source = "source",
              Target = "target", Value = "value", NodeID = "name",
              NodeGroup = "group", LinkGroup = "group",
              fontSize = 12, nodeWidth = 30, 
              height=800, width=1000)
```

The Sankey diagram was created using the very helpful R package [networkD3](https://christophergandrud.github.io/networkD3/)